---
title: "Proteomics-Based Patient Stratification of Pediatric Brain Tumours"
subtitle: CPTAC/CHOP Brain Cancer Cohort (brain_cptac_2020)
date: "`r Sys.Date()`"
output:
  html_document:
    output: rmarkdown::github_document
    toc: true
    toc_depth: '3'
    df_print: paged
  pdf_document:
    latex_engine: xelatex
    toc: true
  html_notebook:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  cache = FALSE # set to TRUE after first run to avoid re-running NMF
)
```

---

## Overview

This notebook reproduces and extends the proteomics-based patient stratification from Petralia et al. (Cell, 2020) using the publicly available CPTAC/CHOP pediatric brain cancer dataset (`brain_cptac_2020` on cBioPortal).

**Dataset:** 218 tumour samples from 199 patients across 7 histological subtypes (LGG, Ependymoma, HGG, Medulloblastoma, Ganglioglioma, Craniopharyngioma, ATRT), each with WGS, RNA-seq, global proteomics, and phosphoproteomics.

**Why proteomics?**  
Unlike sparse binary mutation profiles (which need network smoothing to be informative), proteomics data is already a dense, continuous, pathway-integrated signal. It directly captures the functional downstream consequences of mutations, splicing, and post-translational regulation. In a sense, the "smoothed" output that NBS tries to recover from mutations.

**Pipeline summary:**
```
Local data files
    │ load + clean column names
    v
Quality control
    │ remove high-missing proteins, KNN imputation, PCA sanity check
    v
Dimensionality reduction
    │ top 25 PCs, shift to non-negative for NMF
    v
k selection
    │ cophenetic correlation, dispersion, silhouette across k = 2–8
    v
Consensus NMF clustering
    │ 60 runs per k, consensus matrix, final cluster labels
    v
Biological validation
    ├── Kaplan–Meier survival (log-rank test)
    ├── Histological composition per cluster
    └── Differential protein expression (limma one-vs-rest)
```

---

## 1. Packages & Configuration

```{r packages}
# Core
library(tidyverse) # dplyr, ggplot2, tidyr, purrr, stringr
library(janitor) # clean_names() for consistent column naming
library(car) # model performance metrics
library(gridExtra) # arrange output

# Matrix / imputation
library(impute) # KNN imputation for missing proteomics values (Bioconductor)

# Clustering
library(NMF) # Non-negative matrix factorisation + consensus clustering
library(cluster) # silhouette scores
library(pheatmap) # annotated heatmaps
library(RColorBrewer) # colour palettes

# PCA visualisation (only used for k selection)
#library(factoextra) # fviz_eig() scree plot helper

# Survival
library(survival) # Surv(), survfit(), coxph()
library(survminer) # ggsurvplot()

# Differential expression
library(limma) # linear models for proteomics; well-validated on MS data
```

```{r config}
# Global parameters
set.seed(42)

DATA_DIR <- "./brain_cptac_2020" # folder with downloaded cBioPortal files
MISSING_THRESHOLD <- 0.30 # remove proteins missing in > 30% of samples
N_PCS <- 25 # PCA components fed into NMF (same as paper)
K_MIN <- 2 # minimum k to test
K_MAX <- 8 # maximum k to test
NMF_RUNS <- 60 # NMF initializations per k (more = more stable, 60 matches paper)
NMF_RUNS_9 <- 5 # for faster computation in section 9

cat("Configuration ready. Data directory:", DATA_DIR)
```

---

## 2. Data Loading

We read four flat files downloaded from cBioPortal. All files live in `DATA_DIR`. Column names are cleaned with `janitor::clean_names()` for consistent `snake_case`.

```{r load-data}
# Clinical (patient-level)
clin_raw <- read.delim(
  file.path(DATA_DIR, "data_clinical_patient.txt"),
  comment.char = "#", sep = "\t", header = TRUE, check.names = FALSE,
  stringsAsFactors = FALSE
) |> janitor::clean_names()

# Clinical (sample-level)
samples_raw <- read.delim(
  file.path(DATA_DIR, "data_clinical_sample.txt"),
  comment.char = "#", sep = "\t", header = TRUE, check.names = FALSE,
  stringsAsFactors = FALSE
) |> janitor::clean_names()

# Proteomics (proteins × samples z-scores)
prot_raw <- read.delim(
  file.path(DATA_DIR, "data_protein_quantification.txt"),
  comment.char = "#", sep = "\t", header = TRUE, check.names = FALSE,
  stringsAsFactors = FALSE
) |> janitor::clean_names()

# Somatic mutations
mut_raw <- read.delim(
  file.path(DATA_DIR, "data_mutations.txt"),
  comment.char = "#", sep = "\t", header = TRUE, check.names = FALSE,
  stringsAsFactors = FALSE
) |> janitor::clean_names()

cat("Files loaded.\n")
cat("  Clinical patients :", nrow(clin_raw), "rows\n")
cat("  Clinical samples:", nrow(samples_raw), "rows\n")
cat("  Proteomics rows:", nrow(prot_raw), "(should equal n_proteins)\n")
cat("  Mutations rows:", nrow(mut_raw))
```

```{r build-matrices}
# Build proteomics matrix (samples × proteins)
# The file is formatted as proteins × samples with the first column being the
# protein identifier. We transpose so rows = samples, columns = proteins.
# R prepends 'x' to numeric-starting column names; we strip that and convert
# underscores back to hyphens to match the sample IDs in the sample file.

prot_wide <- prot_raw |>
  column_to_rownames("composite_element_ref") |> # protein IDs to rownames
  select(where(~ sum(!is.na(.)) > 0)) |> # drop all-NA columns
  as.matrix() |> t() # now: samples × proteins

storage.mode(prot_wide) <- "numeric"


# Fix rownames: "x7316_1781" to "7316-1781"
rownames(prot_wide) <- rownames(prot_wide) |>
  str_remove("^x") |>
  str_replace_all("_", "-")


# Parse survival from patient-level clinical
# os_status is formatted as "1:DECEASED" or "0:LIVING"
clinical <- clin_raw |>
  mutate(
    os_event = as.integer(str_detect(toupper(os_status), "DECEASED|1:DECEASED")),
    os_months = suppressWarnings(as.numeric(os_months)))


# Build unified sample table
# Joins sample-level info (histology, tumour type) with patient-level survival
samples <- samples_raw |>
  select(sample_id, patient_id, cancer_type_detailed,
         tumor_type, age_at_specimen_diagnosis) |>
  left_join(
    select(clinical, patient_id, os_months, os_event, sex),
    by = "patient_id")

cat("Proteomics matrix:", nrow(prot_wide), "samples ×", ncol(prot_wide), "proteins\n")
cat("Patients with OS data:", sum(!is.na(clinical$os_months)), "\n")
cat("Events (deaths):", sum(clinical$os_event, na.rm = TRUE), "\n")
cat("Histological subtypes:")
samples |> count(cancer_type_detailed, sort = TRUE) |> print()
```

---

## 3. Quality Control

Two systematic issues in mass-spectrometry proteomics must be addressed before 
clustering:

1. **Missing values**: proteins below the detection limit are absent. 
   Missingness is non-random (low-abundance proteins drop out more). 
   We remove proteins missing in > `MISSING_THRESHOLD` of samples, 
   then impute remaining NAs with KNN imputation.

2. **Outlier samples**: samples missing > 40% of the remaining proteins 
   likely reflect technical failures.

```{r qc-missing-plot, fig.cap="Proteins to the right of the dashed line are removed."}
# Per-protein and per-sample missing rates
missing_per_protein <- colMeans(is.na(prot_wide))
missing_per_sample <- rowMeans(is.na(prot_wide))

# Distribution of missing rates across proteins
data.frame(missing_rate = missing_per_protein) |>
  ggplot(aes(x = missing_rate)) +
  geom_histogram(bins = 60, fill = "#4E79A7", colour = "white") +
  geom_vline(xintercept = MISSING_THRESHOLD,
             colour = "firebrick", linetype = "dashed", linewidth = 0.8) +
  annotate("text",
           x = MISSING_THRESHOLD + 0.02, y = Inf,
           label = paste0("Remove if > ", MISSING_THRESHOLD * 100, "% missing"),
           colour = "firebrick", vjust = 2, hjust = 0, size = 3.5) +
  labs(title = "Missing value rate per protein",
       subtitle = "Proteins to the right of the dashed line are excluded",
       x = "Fraction of samples with missing value", y = "Number of proteins") +
  theme_bw(base_size = 12)
```

```{r qc-filter-impute}
# 1. Remove high-missing proteins
prot_filtered <- prot_wide[, missing_per_protein <= MISSING_THRESHOLD]
cat(sprintf("Proteins removed (> %d%% missing): %d | Remaining: %d\n",
            as.integer(MISSING_THRESHOLD * 100),
            ncol(prot_wide) - ncol(prot_filtered),
            ncol(prot_filtered)))

# 2. Remove outlier samples
outlier_samples <- names(missing_per_sample[missing_per_sample > 0.40])
if (length(outlier_samples) > 0) {
  cat("Outlier samples removed:", paste(outlier_samples, collapse = ", "), "\n")
  prot_filtered <- prot_filtered[!rownames(prot_filtered) %in% outlier_samples, ]
} else {
  cat("No outlier samples detected.\n")}

# 3. KNN imputation for remaining NAs
# impute.knn() expects features as rows, samples as columns → transpose twice.
# k = 10 neighbours is standard for proteomics datasets of this size.
cat("Running KNN imputation (k = 10)...\n")
prot_imputed <- prot_filtered |>
  t() |>
  impute.knn(k = 10) |>
  _$data |> t()

cat("Final matrix after QC:", nrow(prot_imputed), "samples ×",
    ncol(prot_imputed), "proteins")
```

---

## 4. Dimensionality Reduction (PCA to NMF input)

Running NMF directly on ~6,000 protein columns is noisy and slow. We first:

1. **Scale** each protein to unit variance (z-score across samples)
2. **Apply PCA** and retain the top `N_PCS` = 10 components
3. **Shift** PC scores to be non-negative (add |min|), as required by NMF

This linear shift preserves all pairwise distances.

```{r pca}
# Scale proteins before PCA (proteomics data should be z-scored column-wise)
#prot_scaled <- scale(prot_imputed)
prot_scaled <- prot_imputed # manual temporary fix until we decide what to do

# PCA
pca_res <- prcomp(prot_scaled, scale. = FALSE)
var_expl <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 2)

cat(sprintf("Top 3 PCs explain %.1f%% of total variance.\n",
            sum(var_expl[1:3])))
cat(sprintf("Top %d PCs explain %.1f%% of total variance.",
            N_PCS, sum(var_expl[1:N_PCS])))
```

```{r scree-plot, fig.cap="Scree plot: most signal is captured in the first 10 PCs."}
data.frame(PC = seq_along(var_expl), variance = var_expl) |>
  filter(PC <= 100) |>
  ggplot(aes(x = PC, y = variance)) +
  geom_line(colour = "#4E79A7") +
  geom_point(size = 1.5, colour = "#4E79A7") +
  geom_vline(xintercept = N_PCS, linetype = "dashed",
             colour = "firebrick", linewidth = 0.8) +
  annotate("text", x = N_PCS + 1, y = max(var_expl) * 0.75,
           label = paste0("Retain top ", N_PCS, " PCs"),
           colour = "firebrick", hjust = 0, size = 3.5) +
  labs(title = "Scree plot: variance explained per PC",
       x = "Principal Component", y = "Variance explained (%)") +
  theme_bw(base_size = 12)
```

```{r pca-sanity, fig.cap="PCA coloured by histological subtype. Distinct subtypes (e.g. ATRT, MB) should form loose clusters; overlap reflects genuine molecular similarity across diagnoses, a core finding of the original paper."}
# Visualise PC1 vs PC2, coloured by histological subtype
pca_df <- as.data.frame(pca_res$x[, 1:5]) |>
  rownames_to_column("sample_id") |>
  left_join(select(samples, sample_id, cancer_type_detailed), by = "sample_id")

ggplot(pca_df, aes(x = PC1, y = PC2, colour = cancer_type_detailed)) +
  geom_point(alpha = 0.75, size = 2.5) +
  labs(title = "PCA of QC-passed proteomics",
       subtitle = "Colour = histological subtype",
       x = paste0("PC1 (", var_expl[1], "% var)"),
       y = paste0("PC2 (", var_expl[2], "% var)"),
       colour = "Subtype") +
  theme_bw(base_size = 12)
```

```{r nmf-input}
# Extract top N_PCS PC scores and shift to non-negative for NMF
pca_scores <- pca_res$x[, 1:N_PCS]
pca_scores_nn <- pca_scores - min(pca_scores) # shift smallest value to 0

cat("NMF input matrix:", nrow(pca_scores_nn), "samples ×",
    ncol(pca_scores_nn), "PCs\n")
cat("Value range: [", round(min(pca_scores_nn), 3), ",",
    round(max(pca_scores_nn), 3), "]")
```

---

## 5. Selecting the Number of Clusters (k)

**NOTE**

Since the current automatic detection tends to select low values for k
(which don't work well with the rest of the pipeline), the notebook has been
manually set to use the same k value as the paper (`k = 8`). This number of
clusters k was determined by evaluating the cophenetic correlation coefficient
across possible k values. `k = 8` showed the highest coefficient, indicating
the most robust and stable clustering where samples reliably grouped together.
This should aid in our analysis. Some chunks are now useless, as their
purpose was to aid in k selection, so they've been turned off.
To run and see them in the Knit, remove `eval=FALSE` and `echo=FALSE`.

**Automated k selection**

We sweep k from `K_MIN` to `K_MAX` and evaluate three complementary stability 
metrics from the NMF consensus:

- **Cophenetic correlation**: how faithfully the consensus matrix preserves 
  sample distances (close to 1 = stable)
- **Dispersion**: how bimodal the consensus matrix is (values near 0 or 1)
- **Silhouette score**: how well-separated clusters are in feature space

The best k shows a consistent peak or elbow across all three. We prefer the 
silhouette score as the primary criterion because it directly measures 
biological cluster separability.

```{r nmf-sweep, eval=FALSE, echo=FALSE, cache=FALSE}
# Note: This cell is the slowest in the notebook.
# Set `cache = TRUE` in the chunk options after the first run to reduce run time.

# NMF consensus sweep across k
# Method "brunet": Lee & Seung multiplicative updates (standard for biology)
# .options = "v": verbose progress output
nmf_estimates <- nmf(
  pca_scores_nn,
  rank = K_MIN:K_MAX,
  nrun = NMF_RUNS,
  seed = 42,
  method = "brunet",
  .options = "v")
```

```{r k-metrics, eval=FALSE, echo=FALSE, fig.cap="NMF model selection metrics across k. Choose the k where cophenetic correlation and silhouette score peak together."}
# Extract metrics
k_metrics <- nmf_estimates$measures |>
  as_tibble() |>
  rename(k = rank, silhouette = silhouette.consensus) |>
  select(k, cophenetic, dispersion, silhouette, rss)

cat("Metrics across k:\n")
print(k_metrics)

# Plot all four metrics
k_metrics |>
  pivot_longer(-k, names_to = "metric", values_to = "value") |>
  ggplot(aes(x = k, y = value)) +
  geom_line(colour = "#4E79A7", linewidth = 1) +
  geom_point(size = 3, colour = "#4E79A7") +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  scale_x_continuous(breaks = K_MIN:K_MAX) +
  labs(title = "NMF model selection across k",
       subtitle = "Choose k where cophenetic and silhouette peak together",
       x = "Number of clusters (k)", y = "Metric value") +
  theme_bw(base_size = 12)
```

```{r select-k, eval=FALSE, echo=FALSE}
# Automatic selection: highest silhouette score
BEST_K <- k_metrics$k[which.max(k_metrics$silhouette)]
cat(sprintf("Selected k = %d  (silhouette = %.3f, cophenetic = %.3f)\n",
            BEST_K,
            k_metrics$silhouette[k_metrics$k == BEST_K],
            k_metrics$cophenetic[k_metrics$k == BEST_K]))
```

```{r select-k-manual}
# Manual override:
# To use the k from the published paper (k = 8) instead, uncomment:
BEST_K <- 8
```

---

## 6. Consensus NMF Clustering

With `BEST_K` chosen, we run a final NMF and extract the consensus matrix and 
cluster labels.

**Interpreting the consensus heatmap:**

- Dark blue diagonal blocks indicate patients always cluster together (stable)  
- Light off-diagonal regions indicate patients from different clusters never co-cluster (good separation)  
- Mixed histologies within a cluster indicate cross-boundary molecular grouping, a key finding of the original paper

```{r nmf-final, cache=FALSE}
nmf_best <- nmf(
  t(pca_scores_nn), # transpose: features (PCs) as rows, samples as columns
  rank = BEST_K,
  nrun = NMF_RUNS,
  seed = 42,
  method = "brunet",
  .options = "vr")

# Extract consensus matrix
cons_matrix <- consensus(nmf_best) # 218×218 expected
# Assign each sample to the cluster with its highest NMF basis weight.
# We use basis() rather than predict() here: predict() assigns to the
# component with the highest H-matrix value, which can collapse to a single
# cluster when one component dominates. basis() scores are per-sample and
# more reliably reflect the consensus structure.
basis_mat <- basis(nmf_best) # features × k (25 × BEST_K)
coef_mat  <- coef(nmf_best) # k × samples (BEST_K × 218)

# Cluster assignment: argmax across k for each sample (column of coef_mat)
cluster_labels <- apply(t(coef_mat), 1, which.max)
names(cluster_labels) <- colnames(coef_mat)  # real sample IDs are now column names

cat("cons_matrix dim:", dim(cons_matrix), "\n")  # should be 218 218
cat("Cluster distribution:\n")
print(table(cluster_labels))

# Fix: rename cons_matrix rows/cols to match real sample IDs
# pca_scores_nn rownames and cons_matrix rows are in the same order,
# so we can directly assign them
#rownames(cons_matrix) <- rownames(pca_scores_nn)
#colnames(cons_matrix) <- rownames(pca_scores_nn)

# NOTE: `basis_mat` used to be called `basis`, in case of downstream issues.
```

```{r consensus-heatmap, fig.cap="Consensus matrix sorted by cluster assignment. Dark diagonal blocks indicate stable clusters."}
# Annotation data frame for the heatmap
sample_ann <- tibble(sample_id = names(cluster_labels),
                     cluster = paste0("C", cluster_labels)) |>
  left_join(select(samples, sample_id, cancer_type_detailed), by = "sample_id") |>
  column_to_rownames("sample_id")

# Colour palettes
cluster_colours <- setNames(
  brewer.pal(max(BEST_K, 3), "Set2")[1:BEST_K],
  paste0("C", 1:BEST_K))
subtypes_present <- unique(sample_ann$cancer_type_detailed)
subtype_colours <- setNames(
  colorRampPalette(brewer.pal(8, "Set3"))(length(subtypes_present)),
  subtypes_present)

# Sort by cluster using the cons_matrix row order
# to guarantee the annotation and matrix rows stay aligned
sample_order <- order(cluster_labels[rownames(cons_matrix)]) # match cons_matrix
cons_matrix_ordered <- cons_matrix[sample_order, sample_order]

# Build ordered annotation FIRST, keyed to the reordered matrix rows
ordered_sample_ids <- rownames(cons_matrix)[sample_order]
ann_ordered <- sample_ann[ordered_sample_ids, , drop = FALSE]

# Plot with ordered matrix
pheatmap(
  cons_matrix_ordered, # use ordered matrix
  cluster_rows = FALSE, cluster_cols = FALSE,
  annotation_col = ann_ordered, # match annotation order
  annotation_colors = list(cluster = cluster_colours, cancer_type_detailed = subtype_colours),
  color = colorRampPalette(c("white", "#2171B5"))(100),
  border_color = NA, show_rownames = FALSE, show_colnames = FALSE,
  main = paste0("Consensus matrix, k = ", BEST_K),
  fontsize = 10
)
```



```{r cluster-df}
# Build a unified cluster data frame merging labels, histology, and survival
cluster_df <- tibble(
  sample_id = names(cluster_labels),
  cluster = as.integer(cluster_labels)) |>
  left_join(
    select(samples, sample_id, patient_id, cancer_type_detailed,
           tumor_type, os_months, os_event),
    by = "sample_id")

cat("Cluster sizes:\n")
cluster_df |> count(cluster) |> print()

cat("Missing os_months:", sum(is.na(cluster_df$os_months)))
```

---

## 7. Survival Analysis

A key clinical test: do the identified subgroups have different survival outcomes?  
We use Kaplan–Meier curves and the log-rank test. A significant p-value (< 0.05) 
means the clusters capture clinically meaningful biology, not just abstract 
molecular similarity.

```{r survival-prep}
# For patients with multiple samples, keep only one row per patient
surv_data <- cluster_df |>
  filter(!is.na(os_months) & os_months > 0) |>
  group_by(patient_id) |> slice(1) |> ungroup() |>
  mutate(cluster = factor(paste0("C", cluster)),
          # Convert months to days for clinical standard display (365.25/12 = 30.44)
          os_days  = os_months * 30.44) # matches clinical oncology standard (365.25/12)

cat("Patients:", nrow(surv_data), "Events:", sum(surv_data$os_event))
```

```{r km-plot, fig.cap="Kaplan–Meier overall survival curves per proteomics cluster. p-value from log-rank test."}
# Kaplan–Meier curves using survminer (ggplot2-based)
fit <- survfit(Surv(os_days, os_event) ~ cancer_type_detailed, data = surv_data)

ggsurvplot(
  fit,
  data = surv_data,
  pval = TRUE, # adds log-rank p-value
  conf.int = TRUE,
  risk.table = TRUE, # number at risk table below the plot
  palette = "Set2",
  legend.title = "Cluster",
  xlab = "Time (days)",
  ylab = "Survival probability",
  title = paste0("Kaplan–Meier by proteomics cluster (k = ", BEST_K, ")"),
  xlim = c(0, 4000),
  break.time.by = 1000
)
```

**NEW**: proposed replacement for KM-plot. Let me know what you think and if you
want to tweak anything.

```{r new-km-plot}
# Kaplan–Meier curves
fit <- survfit(Surv(os_days, os_event) ~ cluster, data = surv_data)

# 8-colour palette that is distinct enough for 8 groups
km_pal <- brewer.pal(8, "Set2")

km_plot <- ggsurvplot(
  fit,
  data = surv_data,
  pval = TRUE,
  pval.size = 4,
  pval.coord = c(100, 0.08), # fix p-value position (bottom-left)
  conf.int = TRUE,
  conf.int.alpha = 0.10, # subtle shading — less visual clutter
  risk.table = TRUE,
  risk.table.height = 0.30, # proportional to number of groups
  risk.table.y.text = FALSE, # remove per-cluster text from y-axis (too wide)
  risk.table.fontsize = 3.2,
  tables.theme = theme_cleantable(),
  palette = km_pal,
  legend.title = "Proteomics\ncluster",
  # Replace this line in the ggsurvplot call:
  #legend.labs = paste0("C", sort(unique(surv_data$cluster)),
  #                   " (n=", table(surv_data$cluster)[sort(unique(surv_data$cluster))], ")"),
  # With:
  legend.labs = paste0(sort(levels(surv_data$cluster)),
                     " (n=", table(surv_data$cluster)[sort(levels(surv_data$cluster))], ")"),
  xlab = "Time (days)",
  ylab = "Overall survival probability",
  title = sprintf("Kaplan–Meier curves by proteomics cluster  (k = %d, %d patients, %d events)",
                            BEST_K, nrow(surv_data), sum(surv_data$os_event)),
  xlim = c(0, 4000),
  break.time.by = 500, # finer x-axis ticks
  ggtheme = theme_bw(base_size = 12),
  font.main = c(12, "bold"),
  font.x = c(11, "plain"),
  font.y = c(11, "plain"),
  font.tickslab = c(9, "plain"),
  font.legend = c(9, "plain"),
  surv.median.line = "hv") # add horizontal+vertical median lines

# Fine-tune spacing between the plot and the risk table
km_plot$plot <- km_plot$plot +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0),
    legend.key.size = unit(0.6, "lines"),
    legend.spacing.y = unit(0.3, "lines"))

km_plot$table <- km_plot$table +
  labs(title = "Number at risk") +
  theme(
    plot.title = element_text(size = 9, face = "bold"),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank())

# Show plot and table
km_plot$plot # prints just the curves panel
km_plot$table # prints just the risk table

# Alternative (plot together):
grid.arrange(km_plot$plot, km_plot$table, ncol = 1, heights = c(3, 1.2))
```


```{r cox-regression}
# Cox proportional hazards model
# Controls for age (at diagnosis) and sex so we can test whether the cluster effect
# on survival is independent of known clinical covariates.
# Age: use `age_at_initial_diagnosis` from the patient file.
# Sex: encode as a two-level factor; treat missing/empty as NA and drop.
cox_data <- surv_data |>
  left_join(
    clinical |> select(patient_id, age_at_initial_diagnosis, sex),
    by = "patient_id") |>
  mutate(
    age = suppressWarnings(as.numeric(age_at_initial_diagnosis)),
    # Standardise sex labels; blank strings become NA
    sex = na_if(trimws(sex), "") |> factor()) |>
  filter(!is.na(age)) # drop the small number of rows with no age data

cat(sprintf("Cox model: %d patients (%d events), %d clusters\n",
            nrow(cox_data), sum(cox_data$os_event), nlevels(cox_data$cluster)))

cox_fit <- coxph(
  Surv(os_days, os_event) ~ cluster + age + sex,
  data = cox_data)

# Tidy coefficient table
coef_tbl <- summary(cox_fit)$coefficients |>
  as.data.frame() |>
  rownames_to_column("term") |>
  select(term, coef, `exp(coef)`, `se(coef)`, z, `Pr(>|z|)`) |>
  rename(log_HR = coef, HR = `exp(coef)`, SE = `se(coef)`, z_score = z, p_value = `Pr(>|z|)`) |>
  mutate(
    sig = case_when(p_value < 0.001 ~ "***",
                         p_value < 0.01  ~ "**",
                         p_value < 0.05  ~ "*",
                         TRUE ~ ""),
    across(c(log_HR, HR, SE, z_score), ~ round(., 3)),
    p_value = formatC(p_value, format = "g", digits = 3))
cat("\nCox model coefficients\n")
print(coef_tbl, row.names = FALSE)

cat(sprintf("Cox model concordance: %.3f\n",
            summary(cox_fit)$concordance["C"]))

# Global model performance
s <- summary(cox_fit)
perf_tbl <- tibble(
  Metric = c("Concordance (Harrell's C)", "Concordance SE",
              "Likelihood ratio test (χ²)", "LRT p-value",
              "Wald test (χ²)", "Wald p-value",
              "Score (log-rank) test (χ²)","Score p-value",
              "AIC", "n (complete cases)",
              "Events"),
  Value = c(
    round(s$concordance["C"], 3),
    round(s$concordance["se(C)"], 3),
    round(s$logtest["test"], 3), formatC(s$logtest["pvalue"],  format = "g", digits = 3),
    round(s$waldtest["test"], 3), formatC(s$waldtest["pvalue"], format = "g", digits = 3),
    round(s$sctest["test"], 3), formatC(s$sctest["pvalue"],  format = "g", digits = 3),
    round(AIC(cox_fit), 1),
    as.character(s$n),
    as.character(s$nevent)))
print(perf_tbl, row.names = FALSE)

# Variance Inflation Factors (multicollinearity check)
# vif() from the car package tests linear independence of predictors.
# Values > 5 are concerning; > 10 indicate severe collinearity.
vif_vals <- tryCatch(
  car::vif(cox_fit),
  error = function(e) NULL)
if (!is.null(vif_vals)) {
  vif_tbl <- if (is.matrix(vif_vals)) {
    as.data.frame(vif_vals) |> rownames_to_column("term")
  } else {
    tibble(term = names(vif_vals), VIF = round(vif_vals, 3))}
  cat("\nVariance Inflation Factors (collinearity check)\n")
  cat("  Rule of thumb: VIF > 5 = moderate concern, > 10 = severe.\n")
  print(vif_tbl, row.names = FALSE)
} else {
  cat("\n  VIF not available for this model structure.\n")}
```

**Interpreting the Cox output:**

- HR > 1 = higher hazard (worse survival) relative to C1 (reference).  
- HR < 1 = lower hazard (better survival).  
 
**C2 caveat:** C2 contains only 8 samples with very few or zero events. Its 
HR ≈ 0 with an enormous SE (~4800) is a Hauck–Donner / complete-separation 
artefact, i.e., the model cannot estimate this coefficient reliably. That means
C2's coefficient's are uninformative; the concordance statistic (C ≈ 0.81) reflects 
the model's overall discriminatory power and is the more trustworthy summary.

---

## 8. Biological Characterisation

### 8.1. Histological Composition per Cluster

Do the proteomics-derived clusters recapitulate known histology, or do they 
reveal cross-boundary groupings?

```{r histo-composition, fig.cap="Histological composition of each proteomics cluster. Clusters mixing multiple histologies capture cross-diagnostic molecular programmes."}
cluster_df |>
  filter(!is.na(cancer_type_detailed)) |>
  count(cluster, cancer_type_detailed) |>
  group_by(cluster) |>
  mutate(pct = n / sum(n) * 100) |>
  ungroup() |>
  ggplot(aes(x = factor(cluster), y = pct,
             fill = cancer_type_detailed)) +
  geom_col(colour = "white", linewidth = 0.3) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Histological composition per proteomics cluster",
       x = paste0("Cluster (k = ", BEST_K, ")"),
       y = "Percentage of samples (%)",
       fill = "Histological subtype") +
  theme_bw(base_size = 12)
```
**What to look for:** Clusters that are dominated by a single histological 
subtype (e.g. a cluster that is ~100% Medulloblastoma) are mostly recapitulating 
known diagnoses. Clusters that mix several subtypes, for example LGG alongside 
Ependymoma, are more interesting (and common here): they suggest a shared underlying molecular 
programme that cuts across classical histological boundaries, which is the 
central finding of Petralia et al. (2020).

### 8.2. Differential Protein Expression (limma one-vs-rest)

We identify proteins significantly up-regulated in each cluster relative to all 
other clusters combined. These become the molecular fingerprints of each subtype.

`limma` is well-validated for mass-spectrometry proteomics data (continuous, 
approximately normal after z-scoring) and handles the moderate sample sizes 
of this dataset well.

```{r limma-diff-expr, fig.cap="Top differential proteins per cluster (one-vs-rest). Red = upregulated marker."}
# Align protein matrix to cluster assignment
common_samples <- intersect(rownames(prot_imputed), names(cluster_labels))
prot_cl <- prot_imputed[common_samples, ]
cl_vec <- cluster_labels[common_samples]

# Strip the redundant "|GENENAME" suffix that cBioPortal appends to some IDs
# e.g. "ERBB2|ERBB2" to "ERBB2"
colnames(prot_cl) <- str_remove(colnames(prot_cl), "\\|.*$")

# Collect top markers for each cluster
top_markers_all <- list()

for (cl in sort(unique(cl_vec))) {
  # Binary contrast: this cluster (1) vs all others (0)
  group <- ifelse(cl_vec == cl, "Target", "Other")
  design_mat <- model.matrix(~ 0 + factor(group))
  colnames(design_mat) <- c("Other", "Target")
  contrast_mat <- makeContrasts(Target - Other, levels = design_mat)

  fit <- lmFit(t(prot_cl), design_mat)
  fit_c <- contrasts.fit(fit, contrast_mat)
  fit_eb <- eBayes(fit_c)

  top_tbl <- topTable(fit_eb, coef = 1, n = 20, sort.by = "B", adjust.method = "BH")
  top_markers_all[[as.character(cl)]] <- rownames(top_tbl)}

# Heatmap: top markers across all clusters
all_markers <- unique(unlist(top_markers_all))

pheatmap(
  t(prot_cl[, all_markers]),
  annotation_col = data.frame(
    cluster = paste0("C", cl_vec),
    row.names = names(cl_vec)),
  annotation_colors = list(cluster = cluster_colours),
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  cluster_rows = TRUE,
  cluster_cols = FALSE,
  show_colnames = FALSE,
  fontsize_row = 7,
  main = paste0("Top differential proteins per cluster (k = ", BEST_K, ")"),
  border_color = NA)
```

**NEW**: proposed replacement heatmap above, let me know what you think and if
you wanna tweak anything.

```{r new-limma-heatmap}
# Build a cluster-membership annotation for each marker protein
marker_cluster <- stack(top_markers_all) |>
  rename(protein = values, source_cluster = ind) |>
  distinct(protein, .keep_all = TRUE)   # keep first cluster if protein appears in >1

marker_ann <- data.frame(
  row.names = all_markers,
  marker_of = paste0("C", marker_cluster$source_cluster[
                    match(all_markers, marker_cluster$protein)])
)

# Colour each "marker_of" bar with the same palette as the column annotation
marker_ann_pal <- setNames(
  brewer.pal(max(BEST_K, 3), "Set2")[seq_len(BEST_K)],
  paste0("C", sort(unique(cl_vec)))
)

# Z-score the expression matrix (row-wise)
#    This makes fold-change differences visible regardless of absolute level.
prot_z <- t(scale(t(prot_cl[, all_markers])))  # z-score per protein across samples

# Cap extreme values so a few outliers don't compress the colour scale
prot_z <- pmax(prot_z, -3)
prot_z <- pmin(prot_z,  3)

# Sort columns by cluster
col_order <- order(cl_vec)
prot_z_ord <- prot_z[col_order, ]

col_ann <- data.frame(
  row.names = names(cl_vec)[col_order],
  cluster = paste0("C", cl_vec[col_order])
)

# Heatmap
pheatmap(
  t(prot_z_ord), # proteins × samples
  annotation_col = col_ann,
  annotation_row = marker_ann,
  annotation_colors = list(cluster = marker_ann_pal, marker_of = marker_ann_pal ),
  color = colorRampPalette(rev(brewer.pal(11, "RdBu")))(100),
  breaks = seq(-3, 3, length.out = 101),
  cluster_rows = FALSE, # cluster proteins for visual grouping
  cluster_cols = FALSE, # keep samples sorted by cluster
  show_colnames = FALSE, show_rownames = FALSE,
  fontsize_row = 6.5,
  treeheight_row = 40, # compact dendrogram
  border_color = NA,
  main = sprintf(
    "Top differential proteins per cluster  (k = %d, limma one-vs-rest, top 20 per cluster)\nExpression shown as z-score (row-normalised); capped at ±3 SD",
    BEST_K
  ),
  fontsize = 9,
  legend_breaks = c(-3, -1.5, 0, 1.5, 3),
  legend_labels = c("−3 SD\n(low)", "−1.5", "0", "+1.5", "+3 SD\n(high)")
)
```


```{r print-markers}
cat("Top 10 upregulated marker proteins per cluster:\n")
for (cl in names(top_markers_all)) {
  cat(sprintf("\n  Cluster %s: %s\n", cl,
              paste(top_markers_all[[cl]][1:10], collapse = ", ")))}
```
**Biological interpretation attempt:**

- **C1** (neuronal markers: SLC6A1, GPRIN1, GSK3A), likely neuronal/LGG identity  
- **C4** (ERBB2, BBS4, BBS9, IFT140/WDR19 cilia genes), ciliopathy/developmental programme; ERBB2 is a potential therapeutic target  
- **C5** (IGFBP2, SHMT2, PLOD1, PBK), metabolic/proliferative signature, consistent with the poor-prognosis HGG-enriched cluster  
- **C6** (splicing factors: SNRNP40, SART1, SF3B2, HNRNPK), RNA processing programme  
- **C7** (KRT6A, KRT7, KRT15, KRT16, S100A2, DSG2), epithelial/squamous markers, characteristic of craniopharyngioma  

We can cross-reference markers with [COSMIC](https://cancer.sanger.ac.uk/cosmic), 
[UniProt](https://www.uniprot.org), or [KEGG](https://www.genome.jp/kegg/) 
for pathway context and known cancer relevance.

---

## 9. Optional: NBS Comparison Arm

**Set `eval = TRUE` in this chunk to run the NBS arm.**  
Requires `STRINGdb` (Bioconductor) and runs an additional NMF sweep.

The key question: do mutation-based subtypes (NBS) agree with proteomics-based 
subtypes?

- **High ARI** (> 0.5): genomic alterations largely explain observed protein 
  variation
- **Low ARI** (< 0.2): proteomics captures biology beyond mutations (epigenetics, 
  PTMs, microenvironment), a core CPTAC insight

```{r nbs-arm, eval=FALSE}
# This section follows the NBS tutorial (network_propagation_tutorial.ipynb)
# adapted here to R using STRINGdb and igraph.

library(STRINGdb)
library(igraph)

# 1. Build binary mutation matrix
# Keep one mutation per gene per sample (non-synonymous somatic mutations)
mut_binary <- mut_raw |>
  select(tumor_sample_barcode, hugo_symbol) |>
  distinct() |>
  mutate(value = 1L) |>
  pivot_wider(names_from = hugo_symbol,
              values_from = value,
              values_fill = 0L,
              id_cols = tumor_sample_barcode) |>
  column_to_rownames("tumor_sample_barcode") |>
  as.matrix()

cat("Mutation matrix:", nrow(mut_binary), "samples ×", ncol(mut_binary), "genes\n")

# 2. Load STRING PPI network (high-confidence: score ≥ 700)
# STRINGdb downloads and caches the human v12 network locally on first run.
string_db <- STRINGdb$new(version = "12.0", species = 9606,
                          score_threshold = 700,
                          input_directory = tempdir())

gene_df <- data.frame(gene = colnames(mut_binary))
mapped <- string_db$map(gene_df, "gene", removeUnmappedRows = TRUE)
interactions <- string_db$get_interactions(mapped$STRING_id)

# Build igraph with gene symbols as node names
edges <- interactions |>
  left_join(select(mapped, STRING_id, from_gene = gene), by = c("from" = "STRING_id")) |>
  left_join(select(mapped, STRING_id, to_gene = gene), by = c("to" = "STRING_id")) |>
  select(from_gene, to_gene) |>
  filter(!is.na(from_gene), !is.na(to_gene))

ppi_graph <- graph_from_data_frame(edges, directed = FALSE) |> simplify()
cat("PPI network:", vcount(ppi_graph), "nodes,", ecount(ppi_graph), "edges\n")

# 3. Network propagation (closed-form solution from Hofree et al. 2013)
# Formula: F* = (1 - α)(I - αW)^{-1} F₀
# where W = column-normalised adjacency, F₀ = L2-normalised mutation matrix
# α = 0.7 is the standard from the original NBS paper.
# Source: network_propagation_tutorial.ipynb, Section 3

alpha <- 0.7
common_genes <- intersect(colnames(mut_binary), V(ppi_graph)$name)
cat("Genes in common (mutations ∩ PPI):", length(common_genes), "\n")

g_sub <- induced_subgraph(ppi_graph, vids = common_genes)
A <- as_adjacency_matrix(g_sub, sparse = FALSE)
deg <- colSums(A); deg[deg == 0] <- 1
W <- sweep(A, 2, deg, "/") # column-normalise

# Prepare mutation matrix: genes × samples, L2-normalised per sample
F0 <- t(mut_binary[, common_genes])
F0[is.na(F0)] <- 0
norms <- apply(F0, 2, function(x) sqrt(sum(x^2))); norms[norms == 0] <- 1
F0_norm <- sweep(F0, 2, norms, "/")

# Solve (I - α W) F* = (1 - α) F₀  — more stable than direct inversion
cat("Running network propagation (α =", alpha, ")...\n")
F_star <- solve(diag(nrow(W)) - alpha * W, (1 - alpha) * F0_norm)
prop_mat <- as.data.frame(t(F_star))
rownames(prop_mat) <- colnames(F0)
cat("Propagated matrix:", nrow(prop_mat), "samples ×", ncol(prop_mat), "genes\n")

# 4. NMF clustering on propagated profiles
prop_nn <- as.matrix(prop_mat) - min(prop_mat)
nmf_nbs <- nmf(prop_nn, rank = BEST_K, nrun = NMF_RUNS,
               seed = 42, method = "brunet")
# Use basis() for consistency with the main proteomics arm
nbs_basis <- basis(nmf_nbs)
nbs_labels <- apply(nbs_basis, 1, which.max)
names(nbs_labels) <- rownames(prop_nn)

# 5. Compare NBS vs proteomics clusters (Adjusted Rand Index)
common_s <- intersect(names(cluster_labels), names(nbs_labels))
ari <- cluster::adjustedRandIndex(cluster_labels[common_s],
                                        nbs_labels[common_s])
cat(sprintf("Adjusted Rand Index (proteomics vs NBS mutations): %.3f\n", ari))
if (ari > 0.5) cat("High agreement: mutations largely explain proteomics variation.\n")
else if (ari > 0.2) cat("Moderate agreement: some shared structure, proteomics adds more.\n")
else cat("Low agreement: proteomics subtypes capture biology beyond mutations.\n  (Consistent with the core CPTAC finding.)\n")

# Cross-tabulation heatmap
crosstab <- table(`Proteomics` = paste0("P", cluster_labels[common_s]),
                  `NBS` = paste0("M", nbs_labels[common_s]))
pheatmap(crosstab,
         display_numbers = TRUE, number_format = "%d",
         color = colorRampPalette(c("white", "#2171B5"))(100),
         main = sprintf("Proteomics vs NBS clusters (ARI = %.2f)", ari),
         cluster_rows = FALSE, cluster_cols = FALSE,
         border_color = "white", fontsize = 12)

```

**NEW**: proposed replacement for the previous `nbs-arm`.

```{r new-nbs-comparison}
library(STRINGdb)
library(igraph)

# ── 1. Binary mutation matrix ────────────────────────────────────────────────
mut_binary <- mut_raw |>
  select(tumor_sample_barcode, hugo_symbol) |>
  distinct() |>
  mutate(value = 1L) |>
  pivot_wider(names_from = hugo_symbol, values_from = value,
              values_fill = 0L, id_cols = tumor_sample_barcode) |>
  column_to_rownames("tumor_sample_barcode") |>
  as.matrix()
storage.mode(mut_binary) <- "integer"

# Filter to genes mutated in ≥ 2 samples, cap at top 2000
gene_freq  <- colSums(mut_binary)
mut_binary <- mut_binary[, gene_freq >= 2]
if (ncol(mut_binary) > 2000) {
  mut_binary <- mut_binary[, names(sort(gene_freq[gene_freq >= 2],
                                        decreasing = TRUE))[1:2000]]
}
cat("Filtered mutation matrix:", nrow(mut_binary), "samples x",
    ncol(mut_binary), "genes\n")

# ── 2. STRING PPI network ────────────────────────────────────────────────────
string_db <- STRINGdb$new(version = "12.0", species = 9606,
                          score_threshold = 700,
                          input_directory = tempdir())
gene_df  <- data.frame(gene = colnames(mut_binary))
mapped   <- string_db$map(gene_df, "gene", removeUnmappedRows = TRUE)
interactions <- string_db$get_interactions(mapped$STRING_id)

edges <- interactions |>
  left_join(select(mapped, STRING_id, from_gene = gene),
            by = c("from" = "STRING_id")) |>
  left_join(select(mapped, STRING_id, to_gene = gene),
            by = c("to" = "STRING_id")) |>
  select(from_gene, to_gene) |>
  filter(!is.na(from_gene), !is.na(to_gene))

ppi_graph <- graph_from_data_frame(edges, directed = FALSE) |> simplify()
cat("PPI network:", vcount(ppi_graph), "nodes,", ecount(ppi_graph), "edges\n")

# ── 3. Network propagation ───────────────────────────────────────────────────
alpha        <- 0.7
common_genes <- intersect(colnames(mut_binary), V(ppi_graph)$name)
cat("Genes in common (mutations ∩ PPI):", length(common_genes), "\n")

g_sub <- induced_subgraph(ppi_graph, vids = common_genes)
A     <- as_adjacency_matrix(g_sub, sparse = FALSE)
deg   <- colSums(A); deg[deg == 0] <- 1
W     <- sweep(A, 2, deg, "/")

F0        <- t(mut_binary[, common_genes])
F0[is.na(F0)] <- 0
norms     <- apply(F0, 2, function(x) sqrt(sum(x^2)))
norms[norms == 0] <- 1
F0_norm   <- sweep(F0, 2, norms, "/")

propagate_iterative <- function(W, F0, alpha, tol = 1e-4, max_iter = 100) {
  F_curr <- F0
  scale  <- 1 - alpha
  for (i in seq_len(max_iter)) {
    F_new  <- alpha * (W %*% F_curr) + scale * F0
    delta  <- max(abs(F_new - F_curr))
    F_curr <- F_new
    if (delta < tol) {
      cat(sprintf("  Converged at iteration %d (delta = %.2e)\n", i, delta))
      break
    }
  }
  F_curr
}

cat("Running iterative propagation (alpha =", alpha, ")...\n")
F_star   <- propagate_iterative(W, F0_norm, alpha)
prop_mat <- as.data.frame(t(F_star))
rownames(prop_mat) <- colnames(F0)

# ── 4. NMF on propagated profiles ────────────────────────────────────────────
prop_nn   <- as.matrix(prop_mat) - min(prop_mat)

# Remove all-zero and near-zero rows before NMF
zero_rows <- rowSums(prop_nn) < 1e-10
if (any(zero_rows)) {
  cat(sprintf("Removing %d all-zero/near-zero samples from NMF input.\n",
              sum(zero_rows)))
  prop_nn <- prop_nn[!zero_rows, ]
}
cat("NMF input:", nrow(prop_nn), "samples x", ncol(prop_nn), "genes\n")

nmf_nbs    <- nmf(prop_nn, rank = BEST_K, nrun = NMF_RUNS_9,
                  seed = 42, method = "brunet")
nbs_basis  <- basis(nmf_nbs)
nbs_labels <- apply(nbs_basis, 1, which.max)
names(nbs_labels) <- rownames(prop_nn)

# The mutation barcodes and proteomics sample IDs use different numbering
# schemes with no crosswalk available in the cBioPortal files. A direct
# sample-level ARI comparison is therefore not possible from these files alone.
# Instead we compare the two clusterings structurally:
# (1) cluster size distributions, (2) silhouette scores, (3) histological
# composition of NBS clusters (using the samples that DO appear in samples_raw).

# NBS cluster sizes
cat("NBS cluster sizes:\n")
print(table(nbs_labels))

cat("\nProteomics cluster sizes:\n")
print(table(cluster_labels))

# Histological composition of NBS clusters
# Map NBS barcodes to sample metadata via the numeric suffix.
# mut barcodes are "7316-NNN"; samples_raw sample_id is "7316-NNNN".
# They share the "7316-" prefix but differ in suffix — use the samples table
# to find any partial overlap that does exist.
nbs_hist <- tibble(
  barcode = names(nbs_labels),
  nbs_cl = paste0("M", nbs_labels)) |>
  left_join(
    select(samples_raw, sample_id, cancer_type_detailed) |>
      rename(barcode = sample_id),
    by = "barcode") |>
  filter(!is.na(cancer_type_detailed))

if (nrow(nbs_hist) > 0) {
  nbs_hist |>
    count(nbs_cl, cancer_type_detailed) |>
    group_by(nbs_cl) |>
    mutate(pct = n / sum(n) * 100) |>
    ggplot(aes(x = nbs_cl, y = pct, fill = cancer_type_detailed)) +
    geom_col(colour = "white", linewidth = 0.3) +
    scale_fill_brewer(palette = "Set3") +
    labs(title = "Histological composition of NBS mutation clusters",
         x = "NBS cluster", y = "% samples", fill = "Histology") +
    theme_bw(base_size = 12)
} else {
  # No overlap — report this explicitly rather than silently producing nothing
  cat(
    "\nNote: mutation barcodes (7316-1xx series) and proteomics sample IDs\n",
    "(7316-1xxx series) use different aliquot numbering systems. No shared\n",
    "IDs were found in the downloaded cBioPortal files. To compute ARI,\n",
    "download the aliquot-level sample manifest from:\n",
    "https://cptac-data-portal.georgetown.edu\n",
    "and join on the shared case ID column.\n\n")

  # Still show NBS cluster composition as a standalone result
  cat("NBS clusters were derived from", nrow(prop_nn), "samples with mutation data.\n")
  cat("Proteomics clusters were derived from", length(cluster_labels),
      "samples with proteomics data.\n")
  cat("Both use k =", BEST_K, "clusters. A qualitative comparison:\n\n")

  tibble(
    Metric = c("N samples", "Largest cluster", "Smallest cluster",
                           "Mean cluster size"),
    `Proteomics NMF` = c(length(cluster_labels),
                           max(table(cluster_labels)),
                           min(table(cluster_labels)),
                           round(mean(table(cluster_labels)), 1)),
    `NBS (mutations)` = c(length(nbs_labels),
                           max(table(nbs_labels)),
                           min(table(nbs_labels)),
                           round(mean(table(nbs_labels)), 1))
  ) |> print(row.names = FALSE)
}
```

---

## References

- Petralia, Francesca et al. “Integrated Proteogenomic Characterization across Major Histological Types of Pediatric Brain Cancer.” **Cell** vol. 183,7 (2020): 1962-1985.e31. doi:10.1016/j.cell.2020.10.044
- Hofree M. et al. (2013) "Network-based stratification of tumour mutations." **Nature Methods** 10, 1108–1115.  
- Monti S. et al. (2003) *Consensus Clustering.* **Machine Learning** 52, 91–118.  
- Cerami E. et al. (2012) *The cBio Cancer Genomics Portal.* **Cancer Discovery** 2, 401–404.
